TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-09 src/paper.tex
update

The discretization of audio has gained significant attention with the advancement of language models (LMs). This method converts audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performance across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs \cite{dac} in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.

Discretization methods have been explored for tasks such as speech enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}, but their application to target speaker extraction (TSE) has been limited. In this paper, we present a novel approach to target speaker extraction using discrete tokens and language models, named \textbf{TSELM}. Inspired by the blind speech separation networks proposed in DASB \cite{dasb}, our model is structured into three stages: encoding, modeling, and decoding.

In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans. The reference speech is passed directly to the encoder, while for mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN from \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training, and an overview of the model is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results. To the best of our knowledge, this is the first study to explore target speaker extraction using discrete tokens. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.

>>>>

The discretization of audio has gained significant attention with the advancement of language models (LMs). It has been explored in
tasks such as speech enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}.
This method converts audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performances across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs \cite{dac} in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.

The discretization approach for Target Speech Extraction (TSE) has received limited attention in the literature. One of the earliest works to introduce discrete tokens in the context of TSE is presented by \cite{gen_tse}, which investigates the use of self-supervised learning (SSL) models such as HuBERT and vq-wav2vec \cite{}. However, this study does not explore the WavLM SSL model, which has shown superior performance in speech separation tasks. Additionally, it primarily utilizes the output from a single layer of the SSL model for discretization. Furthermore, the evaluation in \cite{gen_tse} focuses solely on speech quality, neglecting key metrics such as speech intelligibility and speaker similarity, both of which are critical for the TSE task. To address these limitations, we propose TSELM, a novel approach that leverages multiple hidden layers of WavLM for discretization and incorporates evaluation metrics for speech intelligibility and speaker similarity.

TSELM has three stages: encoding, modeling and decoding. In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans. The reference speech is passed directly to the encoder, while for mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN from \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training, and an overview of the model is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results. To the best of our knowledge, this is the first study to explore target speaker extraction using discrete tokens. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.
